Inferring generation-interval distributions from contact tracing data: \\ \emph{in prep, PRSB}

### 2019-06-04 doc/network.tex
near complete?

\jd{I want to keep thinking about network choice. Would a random network be useful? Are we careful enough about time windows (exponential phase vs (roughly) full single epidemic?}
\swp{I might have an idea... let me think about this a bit more...}
\begin{figure}[!pbth]
\includegraphics[width=\textwidth]{../fig/local_effect.pdf}
\caption{
\textbf{Spatial effects on realized generation intervals.}
Theoretical distributions and means are shown in color (and are the same in each panel, for reference). Simulated distributions and means are shown in black.
(Left) the intrinsic generation-interval distribution corresponds to all contacts by a focal individual, regardless of whether the individual contacted is susceptible (simulated on a star network).
(Middle) the egocentric generation-interval distribution corresponds to the distribution of all infectious contacts by the focal individual with susceptible individuals, in the case where the focal individual is the only possible infector (simulated on a star network).
(Right) realized generation-interval distributions are shorter than egocentric distributions in general, because contacts can be wasted when susceptibles become infected through other routes (simulated on a homogeneous network).
All figures were generated using 5000 stochastic simulations on a network with 5 nodes (1 infector and 4 susceptibles) with Ebola-like parameters \citep{who2014ebola}:
mean latent period $1/\sigma = 5 \textrm{ days}$ and mean infectious period $1/\gamma = 11.4 \textrm{ days}$. 
Per-pair contact rate $\lambda = 0.25 \textrm{ days}^{-1}$ is chosen to be sufficiently high so that the differences between generation-interval distributions are clear.
}
\label{fig:local}
\end{figure}

@@

\swp{Should we move some of the explanation in this section (pop v.~individual ) to Methods?}

@@

\begin{figure}[!pbth]
\includegraphics[width=\textwidth]{../fig/cmp_reproductive.pdf}
\caption{\textbf{Comparison of estimates of reproductive number based on various methods.}
Using the observed generation-interval distributions without correcting for right-censoring severely underestimates the reproductive number.
Similarly, using intrinsic and egocentric generation-interval distributions without accounting for local spatial effects overestimates the reproductive number.
Both population-level and individual-level methods provide estimates of reproductive number that are consistent with the empirical estimates, which we define as the average number of secondary cases generated by the first 100 infected individuals.
Boxplots are generated using 100 stochastic simulations of the SEIR model on an empirical network using Ebola-like parameters \citep{who2014ebola}: mean latent period $1/\sigma = 5 \textrm{ days}$ and mean infectious period $1/\gamma = 11.4 \textrm{ days}$. 
Per-pair contact rate $\lambda = 0.08 \textrm{ days}^{-1}$ is chosen to be sufficiently high such that differences are clear. \swp{Appendix}
}
\label{fig:cmp}
\end{figure}

@@

To test these methods, we simulate 100 epidemics with Ebola-like parameters on an empirical network \citep{leskovec2016snap} and compare the estimates of reproductive number with empirical reproductive numbers, which we define as the average number of secondary cases generated by the first 100 infected individuals (\fref{cmp}).
As we predicted earlier, calculating reproductive number based on the intrinsic generation-interval distribution overestimates the empirical reproductive number;
estimates based on the egocentric generation-interval distribution (\eref{conditional}) are closer to empirical reproductive numbers but still suffer from overestimation as they do not account for indirect spatial effects. Direct estimates based on the observed generation intervals from contact tracing severely underestimates the empirical estimates.
While both population- and individual-level corrections provide similar estimates to the empirical reproductive number on average,
population-level estimates are more variable as they are more sensitive to outliers in generation intervals and our estimate of the growth rate.
For smaller values of \RR, we expect the differences to become smaller. 
\swp{TODO: In Appendix, we present the same figure using Erlang-distributed latent periods, which better corresponds to Ebola. Overall, our qualitative conclusions do not change.}

@@

\swp{Conclusion?}
\jd{Let me try.}
\swp{How about this?}
Contact tracing provides an effective way of controlling the disease and gathering epidemiological information early in an outbreak.
Out study demonstrates that generation intervals measured through contact tracing capture key characteristics (e.g., contact structure) an outbreak; 
these implicit features allow us to infer the ``correct'' generation-interval distribution, which accurately represents how disease spreads in a population. 
Future studies should carefully consider how underlying epidemiological patterns can be inferred from contact tracing data.
Understanding the role of contact tracing and developing protocols that are informed by the theory will help researchers better prepare for future outbreaks.

@@

\jd{Add flowchart?}

@@

We estimate the exponential growth rate $r$ of an epidemic from daily incidence by modeling the cumulative incidence $c(t)$ with a logistic function \citep{ma2014estimating}: 
\begin{equation}
c(t) = \frac{K}{1 + \left[(K/c_0) - 1\right] e^{-rt}},
\end{equation}
where $K$ is the final size of an epidemic and $c_0$ is the initial number of cases.
Fitting this curve directly to cumulative incidence can lead to overly confident result \citep{king2015avoidable}; instead we fit interval incidence $x(t) = c(t + \Delta t) - c(t)$, where $\Delta t$ is 1 day, to daily incidence, assuming that daily incidence follows a negative binomial distribution. 
This assumption requires us to estimate the dispersion parameter $\theta$ of a negative binomial distribution.
Parameters $r$, $K$, $c_0$, and $\theta$ are estimated by maximum likelihood.

@@

\section{Appendix}

>>>>

\jd{I want to keep thinking about network choice. Would a random network be useful? Are we careful enough about time windows (exponential phase vs (roughly) full single epidemic?}
\swp{Clarified time windows.}
\begin{figure}[!pbth]
\includegraphics[width=\textwidth]{../fig/local_effect.pdf}
\caption{
\textbf{Spatial effects on realized generation intervals.}
Theoretical distributions and means are shown in color (and are the same in each panel, for reference). Simulated distributions and means are shown in black.
(Left) the intrinsic generation-interval distribution corresponds to all contacts by a focal individual, regardless of whether the individual contacted is susceptible (simulated on a star network).
(Middle) the egocentric generation-interval distribution corresponds to the distribution of all infectious contacts by the focal individual with susceptible individuals, in the case where the focal individual is the only possible infector (simulated on a star network).
(Right) realized generation-interval distributions are shorter than egocentric distributions in general, because contacts can be wasted when susceptibles become infected through other routes (simulated on a homogeneous network).
All figures were generated using 5000 stochastic simulations on a network with 5 nodes (1 infector and 4 susceptibles) with Ebola-like parameters \citep{who2014ebola}:
mean latent period $1/\sigma = 5 \textrm{ days}$ and mean infectious period $1/\gamma = 11.4 \textrm{ days}$. 
Per-pair contact rate $\lambda = 0.25 \textrm{ days}^{-1}$ is chosen to be sufficiently high so that the differences between generation-interval distributions are clear.
Each simulation is run until all individuals are either susceptible or has recovered.
}
\label{fig:local}
\end{figure}

@@

@@

\begin{figure}[!pbth]
\includegraphics[width=\textwidth]{../fig/cmp_reproductive.pdf}
\caption{\textbf{Comparison of estimates of reproductive number based on various methods.}
Using the observed generation-interval distributions (based on the first 1000 generation intervals) without correcting for right-censoring severely underestimates the reproductive number.
Similarly, using intrinsic and egocentric generation-interval distributions overestimates the reproductive number because they fail to account for local spatial effects.
Both population-level and individual-level methods provide estimates of reproductive number that are consistent with the empirical estimates, which we define as the average number of secondary cases generated by the first 100 infected individuals.
Boxplots are generated using 100 stochastic simulations of the SEIR model on an empirical network using Ebola-like parameters \citep{who2014ebola}: mean latent period $1/\sigma = 5 \textrm{ days}$ and mean infectious period $1/\gamma = 11.4 \textrm{ days}$. 
Per-pair contact rate $\lambda = 0.08 \textrm{ days}^{-1}$ is chosen to be sufficiently high such that differences are clear.
}
\label{fig:cmp}
\end{figure}

@@

To test these methods, we simulate 100 epidemics with Ebola-like parameters on an empirical network \citep{leskovec2016snap} and compare the estimates of reproductive number with empirical reproductive numbers, which we define as the average number of secondary cases generated by the first 100 infected individuals (\fref{cmp}).
As we predicted earlier, calculating reproductive number based on the intrinsic generation-interval distribution overestimates the empirical reproductive number;
estimates based on the egocentric generation-interval distribution (\eref{conditional}) are closer to empirical reproductive numbers but still suffer from overestimation as they do not account for indirect spatial effects. Direct estimates based on the observed generation intervals from contact tracing severely underestimates the empirical estimates.
While both population- and individual-level corrections provide similar estimates to the empirical reproductive number on average,
population-level estimates are more variable as they are more sensitive to outliers in generation intervals and our estimate of the growth rate.
For smaller values of \RR, we expect the differences to become smaller.
In Appendix, we present the same figure using smaller \RR (Appendix A1) and using Erlang-distributed latent periods (Appendix A2), which better corresponds to Ebola.
Overall, our qualitative conclusions do not change.

@@

\swp{Conclusion?}
\jd{Let me try.}
\swp{How about this?}
Contact tracing provides an effective way of controlling the disease and gathering epidemiological information early in an outbreak.
Out study demonstrates that generation intervals measured through contact tracing can capture key characteristics (e.g., contact structure) an outbreak; 
these implicit features allow us to infer the ``correct'' generation-interval distribution, which accurately describes how disease spreads in a population. 
Future studies should carefully consider how underlying epidemiological patterns can be inferred from contact tracing data.
Understanding the role of contact tracing and developing protocols that are informed by the theory will help researchers better prepare for future outbreaks.

@@

\jd{Add flowchart?}
\swp{Not sure what kind of flowchart you're thinking about.. let me think more...}

@@

We estimate the exponential growth rate $r$ of an epidemic from daily incidence by modeling the cumulative incidence $c(t)$ with a logistic function \citep{ma2014estimating}: 
\begin{equation}
c(t) = \frac{K}{1 + \left[(K/c_0) - 1\right] e^{-rt}},
\end{equation}
where $K$ is the final size of an epidemic and $c_0$ is the initial number of cases.
Fitting this curve directly to cumulative incidence can lead to overly confident result \citep{king2015avoidable}; instead we fit interval incidence $x(t) = c(t + \Delta t) - c(t)$, where $\Delta t$ is 1 day, to daily incidence, assuming that daily incidence follows a negative binomial distribution. 
This assumption requires us to estimate the dispersion parameter $\theta$ of a negative binomial distribution.
Parameters $r$, $K$, $c_0$, and $\theta$ are estimated by maximum likelihood.
Fitting time window is defined from the last trough before the peak of an epidemic to the first day after the peak of an epidemic.
\swp{This is using epigrowthfit. How do I cite it? It's not out yet...}

@@

\pagebreak
\appendix
\section{Appendix}

\subsection{Comparison of estimates of reproductive number -- smaller $\RR$}
\pagebreak

\begin{figure}[!pbth]
\includegraphics[width=\textwidth]{../fig/cmp_reproductive.pdf}
\caption{\textbf{Comparison of estimates of reproductive number based on various methods.}
This figure matches \fref{cmp} in the main text.
A smaller per-pair contact rate ($\lambda = 0.0026 \textrm{ days}^{-1}$) is used to simulate epidemics.
All other parameters are the same as in \fref{cmp}.
}
\label{fig:cmpsmall}
\end{figure}

\subsection{Comparison of estimates of reproductive number -- Erlang distributed latent periods}

\begin{figure}[!pbth]
\includegraphics[width=\textwidth]{../fig/cmp_reproductive.pdf}
\caption{\textbf{Comparison of estimates of reproductive number based on various methods.}
This figure matches \fref{cmp} in the main text.
Erland-distributed latent period ($n_E=2$) is used to simulate epidemics; this assumption better matches the incubation period distribution of the Ebola virus disease than the exponential assumption \citep{who2014ebola}.
All other parameters are the same as in \fref{cmp}.
}
\label{fig:cmpseminir}
\end{figure}

\pagebreak
